Intentional release of native species undermines ecological stability

### 2022-10-02 rmd/supporting_info.Rmd
edit: si

Under each simulation scenario, we ran `r attr(list_param[[1]], "n_sim")` timesteps of community dynamics for $100$ values of release level $R$ (0 to 500 with an equal interval). We introduced 50 individuals of each species and allowed them to grow with no intentional release for the first `r unique(list_sim[[1]]$n_warmup)` timesteps (initialization). After the initialization period, we released $R$ individuals of the enhanced species every timestep over the next `r unique(list_sim[[1]]$n_burnin)` timesteps to reach a new equilibrium with the intentional release (burn-in period). We continued the simulation run with intentional release and saved the last `r unique(list_sim[[1]]$n_timestep)` timesteps, which were used to obtain the following summary statistics of the whole community $\sum_i^S N_i$: the coefficient of variation (CV), the temporal mean ($\mu$), and the SD ($\sigma$).

@@

@@

## Theoretical prediction (two-species community)

```{r fig-2sp, fig.width=9, fig.height=9, fig.cap = paste0("Theoretical predictions in a two-species community. **(A)** Spearman's rank correlations between intentional release and community dynamics in a strong competition scenario ($\\bar{\\alpha}=$ ", max(list_param[[1]]$alpha) ,"). Rows distinguish different summary statistics of community dynamics (top: CV, middle: mean, bottom: SD) while columns distinguish deterministic ($\\sigma_{\\varepsilon}=$ ", min(list_param[[1]]$sd_env) ,") and stochastic simulations ($\\sigma_{\\varepsilon}=$ ", max(list_param[[1]]$sd_env) ,"). Cells indicate 400 combinations of carrying capacity and intrinsic growth rate (20 values each). Colors are proportional to Spearman's rank correlations. Dots indicate parameter combinations that are used in the whole community simulation. **(B)** Examples of community dynamics with and without intentional release. Community dynamics diverge at the broken vertical lines, the timestep at which intentional release begins. Rows distinguish intrinsic growth rates, while columns distinguish deterministic and stochastic simulations. Colors indicate release levels, and line types differ in carrying capacities.")}

>>>>

Under each simulation scenario, we ran `r attr(list_param[[1]], "n_sim")` timesteps of community dynamics for $100$ values of release level $R$ ($0$ to $500$ with an equal interval). We introduced $50$ individuals of each species and allowed them to grow with no intentional release for the first `r unique(list_sim[[1]]$n_warmup)` timesteps (initialization). After the initialization period, we released $R$ individuals of the enhanced species every timestep over the next `r unique(list_sim[[1]]$n_burnin)` timesteps to reach a new equilibrium with the intentional release (burn-in period). We continued the simulation run with intentional release and saved the last `r unique(list_sim[[1]]$n_timestep)` timesteps, which were used to obtain the following summary statistics of the whole community $\sum_i^S N_i$: the coefficient of variation (CV), the temporal mean ($\mu$), and the SD ($\sigma$).

@@

# Supplementary text

## Two-species model

The analysis of a two-species community revealed the distinct parameter space predicting a stabilizing, neutral, or destabilizing effect of intentional release. The left column in **Figure S**\@ref(fig:fig-2sp-w)A) shows the predictions in scenarios with no environmental stochasticity ($\sigma_{\varepsilon} =$ `r min(list_param[[1]]$sd_env)`) and weak competition ($\bar{\alpha}=$ `r min(list_param[[1]]$alpha)`). For lower values of intrinsic growth rate $r_i$ ($<1.5$), intentional release had neutral (with higher $K$) to destabilizing effects (with lower $K$) on community dynamics. In the meantime, stabilization with intentional release occurred for higher values of $r_i$ ($\ge 1.5$) by dampening limit cycles or chaos (**Figure S**\@ref(fig:fig-2sp-s)), which is reflected in decreased SD $\sigma$ (**Figure S**\@ref(fig:fig-2sp-w)). However, the long-term mean of community density $\mu$ consistently decreased with intentional release in almost all cases except for only a few combinations of $r_i$ and $K$ (**Figure S**\@ref(fig:fig-2sp-w)). Environmental stochasticity did not change the qualitative patterns of deterministic simulations but greatly expanded the parameter space with destabilizing effects of intentional release (the right columns in **Figure S**\@ref(fig:fig-2sp-w)**A and B**). Increasing the competition coefficient $\bar{\alpha}$ shifted the region with the destabilizing effects to higher carrying capacities ($\bar{\alpha}=$ `r max(list_param[[1]]$alpha)`, **Figure S**\@ref(fig:fig-2sp-s)); otherwise, general patterns were similar to the weak competition scenario.

@@

## Theoretical prediction (two-species community with weak competition)

```{r fig-2sp-w, fig.width=9, fig.height=9, fig.cap = paste0("Theoretical predictions in a two-species community. **(A)** Spearman's rank correlations between intentional release and community dynamics in a weak competition scenario ($\\bar{\\alpha}=$ ", min(list_param[[1]]$alpha) ,"). Rows distinguish different summary statistics of community dynamics (top: CV, middle: mean, bottom: SD) while columns distinguish deterministic ($\\sigma_{\\varepsilon}=$ ", min(list_param[[1]]$sd_env) ,") and stochastic simulations ($\\sigma_{\\varepsilon}=$ ", max(list_param[[1]]$sd_env) ,"). Cells indicate 400 combinations of carrying capacity and intrinsic growth rate (20 values each). Colors are proportional to Spearman's rank correlations. Dots indicate parameter combinations that are used in the whole community simulation. **(B)** Examples of community dynamics with and without intentional release. Community dynamics diverge at the broken vertical lines, the timestep at which intentional release begins. Rows distinguish intrinsic growth rates, while columns distinguish deterministic and stochastic simulations. Colors indicate release levels, and line types differ in carrying capacities.")}

source(here::here("code/figure_theory_2sp.R"))
knitr::include_graphics(here::here("output/figure_2sp_model_weak.pdf"))
```

## Theoretical prediction (two-species community with strong competition)

```{r fig-2sp-s, fig.width=9, fig.height=9, fig.cap = paste0("Theoretical predictions in a two-species community. **(A)** Spearman's rank correlations between intentional release and community dynamics in a strong competition scenario ($\\bar{\\alpha}=$ ", max(list_param[[1]]$alpha) ,"). Rows distinguish different summary statistics of community dynamics (top: CV, middle: mean, bottom: SD) while columns distinguish deterministic ($\\sigma_{\\varepsilon}=$ ", min(list_param[[1]]$sd_env) ,") and stochastic simulations ($\\sigma_{\\varepsilon}=$ ", max(list_param[[1]]$sd_env) ,"). Cells indicate 400 combinations of carrying capacity and intrinsic growth rate (20 values each). Colors are proportional to Spearman's rank correlations. Dots indicate parameter combinations that are used in the whole community simulation. **(B)** Examples of community dynamics with and without intentional release. Community dynamics diverge at the broken vertical lines, the timestep at which intentional release begins. Rows distinguish intrinsic growth rates, while columns distinguish deterministic and stochastic simulations. Colors indicate release levels, and line types differ in carrying capacities.")}
